Preparing for Potential Closure of European Airspaces due to Reentering Space Objects

### 2025-05-02 paper.qmd
better wording

author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2810-1404
    email: rainer.koelle@eurocontrol.int
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2187-1924
    email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium

@@

# Abstract
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice.  
We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space it is clear a re-evaluation of the risks of
collision and casualties due to uncontrolled debris re-entries is needed.
Overseeing authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

Surging number of flights in Europe [@forecastUpdate2025] and in the
world, but more importantly under development/deployment mega-constellations
of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
of orbiting objects with a short operational life. Some of these will 
inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].

This paper develops a practical and straightforward pipeline to rapidly
assess the risk from uncontrolled re-entry of space debris for the European
region.
Apart from the general risk assessment of potential re-entries, 
a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus is analysed.

@@

# ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
megas |> 
  gt() |> 
   cols_label(
    constellation ~ "Constellation",
    state ~ "Country",
    launched ~ "Launched",
    active ~ "Active",
    planned ~ "Planned",
    first_launch ~ "First launch"
  ) |> 
   sub_missing(
    columns = "first_launch",
    missing_text = "-"
  ) |> 
  opt_table_font(size = "10px")

@@

We follow the same mathematical approach of 
[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
flight trajectories.

The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
This work will use H3 cells at resolution 3 which corresponds to hexagons of an
average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).

@@

The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.

This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
provides a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

Next sections will develop the mathematical background for the evaluation of
collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
Expected outcomes such as cells at zero density for war or no-fly zones
(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted.

@@

Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
:::

## Impact probability for a single flight and single debris
The probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

Using H3 to discretise the area of interest $a$ using H3 cells gives

$$
p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h).
$$

Where for each cell $h$ the following applies

$$
p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h).
$$

@@

It is assumed the aircraft impact area to be much larger than the debris
object area so as to neglect the debris impact area.
The geometry of the aircraft is simplified and modelled as
a rectangular cuboid with dimensions of wingspan, length and height. 
Now, if the aircraft were stationary, the area exposed to a debris impact would
be the top view, while if the debris were stationary it would be the front view.
In reality an aircraft en-route is moving (mainly) horizontally and the debris 
vertically [^balistic_descent], which which allows to express the effective
exposed area, $\varepsilon_k$, as

@@

In the calculations the following is assumed to be true ($t$ subscript for aircraft type)

@@

[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
path angle of a re-entering object at 40,000 feet is -86.6 degrees, which
corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than 50 lbs/ft2, the vertical descent assumption is a very
accurate approximation.* [@patera2008]

@@

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to re-entry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta$, of the re-entry point from the ascending node, 
see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.

@@

If we consider historical debris re-entry data, we can numerically derive
the debris impact probability as a weighting function by summing up and averaging
all single densities:

@@

From [@gcat] we extracted 1968 debris re-entries during 25 calendar years from 
1st Jan 2000 till 31st Dec 2024.
The result of combining all the U-shaped densities can be seen in @fig-weighting.

@@

Weighting function derived from 25 years of re-entries as an
empirical approximation of $\sigma(\varphi)$. 
:::

@@

## Hourly impact probability per H3 cell
By combining the hourly aircraft density (per aircraft type) with the relevant
weightings per H3 cell, we can numerically compute the hourly impact probability
for each hexagonal cell as:

$$
 E_h =  \sum_t W(h) \cdot \rho_{t}(h) \cdot \varepsilon_t
$$

where $E_h$ is the collision expectation in the hexagon $h$,
$W(h)$ is the median value of the weighting function over the area of hexagon $h$, $t$ is the \acr{ICAO} aircraft type,
$\rho_{t}(h)$ is the traffic density for type $t$ in cell $h$ and
$\varepsilon_T$ is effective exposed area for aircraft of type $t$.

@@

```{r collision}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/example-image.png")
```

Collision risk probability. 
:::

## Use Case: KOSMOS 482 Descent Craft re-entry

In few days (at the time of writing this paper) at 6h May 10, 2025
($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
Earth's atmosphere [@kosmos482re-entry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission
[@SpaceReviewKosmos] of \qty{480}{kg} weight and circa \qty{1}{m} diameter.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \qty{467}{degreesCelsius} (\qty{872}{degreesFarenheit})).
So it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

With what developed in this paper, we can derive the casuality risk and
determine the maximum value using data for the busiest day in 
the busiest 2024 week.

Once the re-entry track is known more or less accurately, the relevant
H3 cells can be

@@

Casuality for KOSMOS 482 Descent Craft re-entry. 
:::

@@

We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

Future work will explore the impact of megacostellations and the risk for
European airspace.

# Appendix {.appendix}
# Orbital Elements {#sec-orbitals }

@@

The angular position, $\theta$, of the re-entry point from the ascending node
in the orbit plane is $\theta = \omega + \nu$.
It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or

>>>>

author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium

@@

# Abstract
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
National and international authorities need to balance safety and economic
impact of adopted measures should a reentry event be announced, usually on short
notice.  
We employ hourly and daily spatial density of flight calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned over Uber's H3 gridding system.  
Impact risk probabilities are then calculated based on previous theoretical
work but using these flights trajectories instead of proxies such as
population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide risk impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

The ever-growing forecast number of flights, see @fig-forecast, and the modern
race to commercialisation of space call for a reevaluation of the risk of
collision and casuality due to uncontrolled debris reentries.
Oversighting authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimize casualties and economic
impact.

Surging number of flights in Europe [@forecastUpdate2025] and in the
world, but more importantly under development/deployment megaconstellations
of satellites on \acr{LEO} such as already active Starlink and OneWeb or
planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
of orbiting objects with a short operational life some of which will 
inevitably reenter Earth atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].

This paper develops a practical and straightforward pipeline to rapidly
assess the risk from uncontrolled reentry of space debris for the European
region.
Apart from the general risk assessment of potential reentries, we analyze
a soon to happen reentry of debris from a 1972 failed USSR mission to Venus.

@@

# ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
megas |> 
  gt() |> 
   cols_label(
    constellation ~ "Constellation",
    launched ~ "Launched",
    active ~ "Active",
    planned ~ "Planned",
    first_launch ~ "First launch"
  ) |> 
   sub_missing(
    columns = "first_launch",
    missing_text = "-"
  ) |> 
  opt_table_font(size = "10px")

@@

We follow the same mathematical approach of 
[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use flown 
trajectories.

We base our hourly flight density estimation on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1] and binned on Uber's H3 gridding system
[@uber-h3] for fast processing.  
This work will use H3 cells at resolution 3 which corresponds to hexagons of
average area $\qty{12393}{km^2}$ and average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
box of 43 EUROCONTROL Member States FIR's (blue shaded polygon).

@@

We characterise densities on an hourly basis for the days of the 2023 and 2024
busiest \acr{IATA} summer ISO week.

We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
provide a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

Next sections will develop the mathematcal background for the evaluation of
collision and casuality risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for 2024-07-05.
Expected outcomes such as cells at zero density for war or no-fly zones
(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted.

@@

Flight density in the EUROCONTROL Area on 2024-07-05, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
:::

## Impact Probability for Single Flight, Single Debris
The probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

If we discretize the area of interest $a$ using H3 cells,
we have

$$
p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h)
$$

So for cell $h$ we have

$$
p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h)
$$

@@

We assume the aircraft impact area to be much larger than the debris
object area so as to neglet the debris impact area.
We can also simplify the geometry of the aircraft and model it as
a rectangular solid with dimesions of wingspan ($W$), length ($L$) and
height ($H$). 
Now, if the aircraft were stationary, the area exposed to a debris impact would
be the top view, while if the debris were stationary it would be the front view.
In reality the aircraft is moving (manly) horizontally and the debris 
vertically [^balistic_descent], which which allows to express the effective
exposed area, $\varepsilon_k$, as

@@

In our calculations we will use ($t$ subscript for aircraft type)

@@

[^balistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
the flight path angle of a reentering object at \qty{40000}{ft} is 
\ang{-86.6}, which corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
accurate approximation.* [@patera2008]

@@

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to reentry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta$, of the reentry point from the ascending node, 
see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.

@@

If we consider historical uncontrolled debris reentry data, we can numerically
derive the debris impact probability as a weighting function by summing up and
averaging all single densities:

@@

From [@gcat] we extracted 1968 debris reentries during 25 calendar years from 
1st Jan 2000 till 31st Dec 2024.
The result of combining all the U-shaped densities can be seen in @fig-weighting.

@@

Weighting function derived from 25 years of reentries as an
empirical approximation of $\sigma(\varphi)$. 
:::

@@

## Hourly impact probability per H3 cell

When combining the hourly ($H$) aircraft density (per aircraft type) with the
relevant weightings per H3 cell, $h$, we can numerically compute the hourly
impact probability for each hexagonal cell as:

$$
 E_{h, H} =  \sum_t W(h) \cdot \rho_{t,H}(h) \cdot \varepsilon_t
$$

where $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
$W(h)$ is the median value of the weighting function over the area of hexagon
$h$,
$\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
cell $h$ at hout $H$ and
$\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@@

```{r collision}
#| out-width: "100%"
#| echo: false
knitr::include_graphics(here::here("media", "figures", "collision_expectation_2024-07-05_3.png"))
```

Averaged hourly collision expectation on 2024-07-05. 
:::

## Use Case: KOSMOS 482 Descent Craft reentry

In few days (at the time of writing of this paper) at 6h May 10, 2025
($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
Earth's atmosphere [@kosmos482reentry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission
[@SpaceReviewKosmos] of \qty{480}{kg} weight and approximately
\qty{1}{m} diameter.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

Knowing that the debris inclination is $\ang{51.7}$,
see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a
proxy for the traffic density on the day of reentry.

Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
can be identified to retrieve the values of collision expectancy.

@@

Casuality for KOSMOS 482 Descent Craft reentry. 
:::

@@

We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled reentry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

Future work will explore the impact of megacostellations and flight growth
forecast and what would happen in 20, 40, ..., 100 years in varios scenarios
of uncontrolled  reentries and the risk for European airspace.

# Appendix: Orbital Elements {#sec-orbitals}

@@

The angular position, $\theta$, of the reentry point from the ascending node
in the orbit plane is $\theta = \omega + \nu$.
It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or
